---
title: "Demographic and epidemiological characteristics of pregnant and postpartum women who died from Severe Acute Respiratory Syndrome in Brazil: a comparison between COVID-19 and nonspecific etiologic causes"
author: 'Codes and outputs'
date: "10/01/2021"
output:   
  pdf_document:
    keep_tex: yes
  word_document: default
  html_document:
    df_print: paged
    self_contained: no
---

# Description
 
This file presents the documentation of the analysis of article "Demographic and epidemiological characteristics of pregnant and postpartum women who died from Severe Acute Respiratory Syndrome in Brazil: a comparison between COVID-19 and nonspecific etiologic causes".

# R packages used, functions and dataset import


The data are analyzed using the free-software R (https://www.R-project.org) in version 4.0.3. Next, we present and load the libraries used in the data analysis process.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r pacotes, echo=TRUE, message=FALSE, warning =FALSE,error=FALSE, results='hide'}
#load packages
loadlibrary <- function(x) {
  if (!require(x, character.only = TRUE)) {
    install.packages(x, dependencies = T)
    if (!require(x, character.only = TRUE))
      stop("Package not found")
  }
}

packages <-
  c(
    "dplyr",
    "lubridate",
    "readr",
    "readxl",
    "ggplot2",
    "kableExtra",
    "tables",
    "questionr",
    "car",
    "data.table",
    "magrittr",
    "tidyverse",
    "readxl",
    "summarytools",
    "modelsummary",
    "RColorBrewer",
    "zoo",
    "grid",
    "gridExtra",
    "cowplot",
    "effectsize",
    "rcompanion",
    "DescTools"
  )
lapply(packages, loadlibrary)
```

One can see below the functions that will be used in the data analysis.
```{r,echo=TRUE, eval=TRUE, message=FALSE,warning =FALSE,error=FALSE,results='hide'}
#functions for summary measures
media <- function(x)
  mean(x, na.rm = TRUE)
mediana <- function(x)
  median(x, na.rm = TRUE)
DP <- function(x)
  sd(x, na.rm = TRUE)
minimo <- function(x)
  base::min(x, na.rm = TRUE)
maximo <- function(x)
  base::max(x, na.rm = TRUE)
q25 <- function(x)
  stats::quantile(x, p = 0.25, na.rm = TRUE)
q75 <- function(x)
  stats::quantile(x, p = 0.75, na.rm = TRUE)
IQR <- function(x)
  round(q75(x) - q25(x), 2)
n <- function(x)
  sum(!is.na(x))
```


The Influenza Epidemiological
Surveillance Information System, SIVEP-Gripe (Sistema de Informação de Vigilância Epidemiológica da Gripe), is a
nationwide surveillance database used to monitor severe acute respiratory infections in Brazil. 

Notification is mandatory for Influenza Syndrome (characterized by at least two of the following signs and symptoms: fever, even if referred, chills, sore throat, headache, cough, runny nose, olfactory or taste disorders) and who has dyspnea/respiratory discomfort or persistent pressure in the chest or O2 saturation less than 95\% in room air or bluish color of the lips or face. Asymptomatic individuals with laboratory confirmation by molecular biology or immunological examination for COVID-19 infection are also reported.

For notifications in Sivep-Gripe, hospitalized cases in both public and private hospitals and all deaths due to severe acute respiratory infections regardless of hospitalization must be considered.

The search was limited to the first notified case of COVID-19 in February 2020 until the 15th epidemiological week of 2021 (up to April 17, 2021). The datasets were obtained on May 5, 2021, on the site https://opendatasus.saude.gov.br/dataset. The first period (8th to 53rd epidemiological week of 2020) and the second period (1st to 15th epidemiological week of 2021) datasets can be obtained at https://drive.google.com/file/d/1jts4h0ovdwFh86SdKyslMLSG9rOy3UjX/view?usp=sharing and at https://drive.google.com/file/d/1gQSy_dcUkd1UrDEcsrDbyGl4gEvcI8z_/view?usp=sharing, respectively. The data are loaded below:

```{r,echo=FALSE, eval=TRUE, message=FALSE,warning =FALSE,error=FALSE,results='hide'}
 memory.limit(999999)
```


```{r,echo=TRUE,message=FALSE,warning =FALSE,error=FALSE,results='hide'}
#loading the datasets
#2021
data_2021 <- read_delim(
  "INFLUD21-03-05-2021.csv",
  ";",
  escape_double = FALSE,
  locale = locale(encoding = "ISO-8859-2"),
  trim_ws = TRUE
)

#2020
data_2020 <- read_delim(
  "INFLUD-03-05-2021.csv",
  ";",
  escape_double = FALSE,
  locale = locale(encoding = "ISO-8859-2"),
  trim_ws = TRUE
)

sem <- 15 #limit of epidemiological week of 2021

#### Concatenating 2020 and 2021 data ##############
data_all <- rbind(data_2020, data_2021)

# Creating the case year variable
data_all <-  data_all %>%
  dplyr::mutate(
    dt_sint = as.Date(DT_SIN_PRI, format = "%d/%m/%Y"),
    year_case = lubridate::year(dt_sint)
  )

```

There are `r dim(data_all)[1]` cases in the complete dataset.  The case selection is presented in the following according to the flowchart presented in the article. 

# Case selection and data treatment

The first filter consists of selecting the cases from 8th epidemiological week of 2020 to 15th epidemiological week of 2021. 

```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
#Cases from the 8th epidemiological week of 2020
data1 <- data_all %>% 
  filter((year_case == 2020 & SEM_PRI >= 8) | year_case == 2021)
```

There are `r dim(data1[data1$year_case==2021  & data1$SEM_PRI==53, ])[1]` cases in 2021 in epidemiological week 53 of 2020. These are cases from the first two days of 2021, which are still part of the last epidemiological week of 2020 (http://portalsinan.saude.gov.br/calendario-epidemiologico?layout=edit&id=168). However, these cases belong to the 53rd week of 2020 and we corrected as follows:

```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
data2 <- data1 %>% 
  mutate(year_case = ifelse(year_case == 2021 & SEM_PRI ==53, 2020, year_case)) %>% 
  filter(year_case == 2020 | (year_case == 2021 & SEM_PRI <= sem)) 
```

There are `r dim(data2)[1]` observations.

The next selection consists of selecting females cases:
```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
#filtering F cases
data3 <- filter(data2, CS_SEXO == "F")
```

There are `r dim(data3)[1]` observations.

The next step is to identify pregnant and postpartum people (variable `classi_gesta_puerp`) and then select only those cases.

```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
#Creating the classification variable if pregnant, postpartum and not (neither pregnant nor postpartum)
data3 <- data3 %>%
  mutate(
    classi_gesta_puerp = case_when(
      CS_GESTANT == 1  ~ "1tri", #1st trimester
      CS_GESTANT == 2  ~ "2tri", #2st trimester
      CS_GESTANT == 3  ~ "3tri", #3st trimester
      CS_GESTANT == 4  ~ "GA_ig", #ignored gestational Age
      CS_GESTANT == 5 &
        PUERPERA == 1 ~ "puerp", #puerperium
      CS_GESTANT == 9 & PUERPERA == 1 ~ "puerp",  #puerperium
      TRUE ~ "no" #neither pregnant nor postpartum 
    )
  )

#Filtering only pregnant and postpartum women
data4 <- data3 %>%
  filter(classi_gesta_puerp != "no")
```

There are `r dim(data4)[1]` observations.

The next selection consists of selecting women between 10 and 55 years old. 

```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
# Filtering ages between 10 and 55 years old
data5 <- data4 %>%
  filter(TP_IDADE == 3 | (TP_IDADE != 3 & NU_IDADE_N >=8) #NU_IDADE_N is the age variable
         ) %>% 
  filter(NU_IDADE_N > 9 & NU_IDADE_N <= 55)
```

There are `r dim(data5)[1]` observations.


Now we are going to select only the cases confirmed by COVID-19 (CLASSI_FIN = 5) or unspecified (CLASSI_FIN = 4). 

```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
#Filtering only covid or unspecified cases
data6 <- data5 %>% 
    filter(CLASSI_FIN == 5 | CLASSI_FIN ==4)
```

```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
freq(data6$CLASSI_FIN)
```

There are `r dim(data6)[1]` observations.

We are now going to select only the death cases. The variable is `EVOLUCAO`, with 1 - cure, 2 - death by SARS, 3 - death by other cause. 

```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
data6 <- data6 %>% 
  mutate(death = case_when(
    EVOLUCAO == 1 ~ "cure", 
    EVOLUCAO == 2 ~ "death", 
    EVOLUCAO == 3 ~ "death", 
    TRUE ~ NA_character_
  ))


data7 <- data6 %>% 
  filter((EVOLUCAO == 2 | EVOLUCAO == 3) & !is.na(EVOLUCAO))
```

There are `r dim(data7)[1]` observations.


Now labeling group variable:

```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
data7 <- data7 %>% 
  mutate(group = case_when(
    CLASSI_FIN ==5 ~ "covid-19", 
    TRUE ~ "unspecified"
  ))
data7$group <- factor(data7$group, 
                     levels = c("unspecified","covid-19"))

```

```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
with(data7, freq(group))
```

# Epidemiologic characteristics 


```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
# Ethnicity
data7 <-  data7 %>%
  mutate(
    ethnicity = case_when(
      CS_RACA == 1 ~ "white",
      CS_RACA == 2 ~ "black",
      CS_RACA == 3 ~ "yellow",
      CS_RACA == 4 ~ "brown",
      CS_RACA == 5 ~ "indigenous",
      TRUE ~ NA_character_
    )
  )

# Education
data7 <-  data7 %>%
  mutate(education = case_when(CS_ESCOL_N == 0 ~ "no education",
                             CS_ESCOL_N == 1 | CS_ESCOL_N == 2 ~ "primary education",
                             CS_ESCOL_N == 3 ~ "secundary education",
                             CS_ESCOL_N == 4 ~ "higher education",
                               TRUE ~ NA_character_))

data7$education <- factor(data7$education, 
                     levels = c("no education","primary education", 
                                "secundary education", "higher education"))

# Age group
data7 <-  data7 %>%
  mutate(
    age_group = case_when(
      NU_IDADE_N <= 19 ~ "<20",
      NU_IDADE_N >= 20
      & NU_IDADE_N <= 34 ~ "20-34",
      NU_IDADE_N >= 35 ~ ">=35",
      TRUE ~ NA_character_
    )
  )
data7$age_group <-
  factor(data7$age_group, levels = c("<20", "20-34", ">=35"))


# Residence area
data7 <-  data7 %>%
  mutate(zone = case_when(CS_ZONA ==1  ~ "urban",
                             CS_ZONA == 2 ~ "rural",
                             CS_ZONA == 3 ~ "periurban",
                               TRUE ~ NA_character_))
```

## Ethnicity

```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
with(data7, ctable(ethnicity, group,  prop = "c", useNA = "no", chisq = FALSE))
```


```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
with(data7, fisher.test(ethnicity, group))
```


## Education

```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
with(data7, ctable(education, group, prop = "c", useNA = "no", chisq = FALSE))
```

```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
with(data7, fisher.test(education, group))
```

## Age 

```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
datasummary((group) ~  NU_IDADE_N*(n+media+DP+mediana+q25+q75+IQR),
            data = data7, output = 'markdown')
```

```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
#t test
t.test(NU_IDADE_N ~ group, data = data7)
```

```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
#effect size
c_cohen <- cohens_d(NU_IDADE_N ~ as.factor(group),data=data7)
c_cohen
interpret_d(c_cohen$Cohens_d,rules="cohen1988")
```

## Age group

```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
with(data7, ctable(age_group, group, prop = "c", useNA = "no", chisq = FALSE))
```

```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
with(data7, fisher.test(age_group, group))
```

## Residence area

```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
with(data7, ctable(zone, group, prop = "c", useNA = "no", chisq = FALSE))
```

```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
with(data7, fisher.test(zone, group))
```

## Gestational moment

```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
with(data7, ctable(classi_gesta_puerp, group, prop = "c", useNA = "no", chisq = TRUE))
```


# Comorbities

```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
#Cardiac
data7 <-  data7 %>%
  mutate(cardiac = case_when(CARDIOPATI == 1 ~ "yes",
                             CARDIOPATI == 2 ~ "no",
                             TRUE ~ NA_character_))

#Hematologic
data7 <-  data7 %>%
  mutate(hematologic = case_when(HEMATOLOGI == 1 ~ "yes",
                                 HEMATOLOGI == 2 ~ "no",
                                 TRUE ~ NA_character_))

#Hepatic
data7 <-  data7 %>%
  mutate(hepatic = case_when(HEPATICA == 1 ~ "yes",
                             HEPATICA == 2 ~ "no",
                             TRUE ~ NA_character_))

#Asthma
data7 <-  data7 %>%
  mutate(asthma = case_when(ASMA == 1 ~ "yes",
                            ASMA == 2 ~ "no",
                            TRUE ~ NA_character_))

#Diabetes
data7 <-  data7 %>%
  mutate(diabetes = case_when(DIABETES == 1 ~ "yes",
                              DIABETES == 2 ~ "no",
                              TRUE ~ NA_character_))

#Neurologic
data7 <-  data7 %>%
  mutate(neurologic = case_when(NEUROLOGIC == 1 ~ "yes",
                                NEUROLOGIC == 2 ~ "no",
                                TRUE ~ NA_character_))

#Pneumologic
data7 <-  data7 %>%
  mutate(pneumologic = case_when(PNEUMOPATI == 1 ~ "yes",
                                 PNEUMOPATI == 2 ~ "no",
                                 TRUE ~ NA_character_))

#Imunossupression
data7 <-  data7 %>%
  mutate(imuno = case_when(IMUNODEPRE == 1 ~ "yes",
                           IMUNODEPRE == 2 ~ "no",
                           TRUE ~ NA_character_))

#Renal
data7 <-  data7 %>%
  mutate(renal = case_when(RENAL == 1 ~ "yes",
                           RENAL == 2 ~ "no",
                           TRUE ~ NA_character_))

#Obesity
data7 <-  data7 %>%
  mutate(obesity = case_when(OBESIDADE == 1 ~ "yes",
                             OBESIDADE == 2 ~ "no",
                             TRUE ~ NA_character_))

#Any comorbidity

df <- data7 %>% 
  select(cardiac,obesity,hematologic,hepatic,asthma,diabetes,neurologic,pneumologic,imuno,renal)

#if all comorbities in df are NA (not available), return NA.
soma <- function(x){
  if (sum(is.na(x))==10)
    return(NA_character_)
  else
    return(sum(!is.na(x) & x=="yes")) 
}
data7$qt_comorb_aux <- apply(df,1,soma)

data7 <- data7 %>%
  mutate(comorbidity = case_when(qt_comorb_aux >= 1 ~ "yes",
                                 qt_comorb_aux == 0 ~ "no",
                                 TRUE ~ NA_character_))
```

## Cardiac

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
with(data7, ctable(group, cardiac, prop = "r", useNA = "no", chisq = TRUE, OR = TRUE))
```

##  Hematologic

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
with(data7, ctable(group, hematologic, prop = "r", useNA = "no", chisq = FALSE, OR = TRUE))
```

```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
with(data7, fisher.test(hematologic, group))
```

##  Diabetes

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
ctable(data7$group, data7$diabetes, chisq=TRUE, prop="r",  useNA = "no", OR = TRUE)
```

##  Obesity

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
ctable(data7$group, data7$obesity, chisq=TRUE, prop="r",  useNA = "no", OR = TRUE)
```

##  Asthma

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
ctable(data7$group, data7$asthma, chisq=TRUE, prop="r",  useNA = "no", OR = TRUE)
```

##  Hepatic

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
ctable(data7$group, data7$hepatic, chisq=FALSE, prop="r",  useNA = "no", OR = TRUE)
```

```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
with(data7, fisher.test(hepatic, group))
```

##  Neurologic

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
ctable(data7$group, data7$neurologic, chisq=FALSE, prop="r",  useNA = "no", OR = TRUE)
```

```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
with(data7, fisher.test(neurologic, group))
```

##  Pneumologic

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
ctable(data7$group, data7$pneumologic, chisq=TRUE, prop="r",  useNA = "no", OR = TRUE)
```

##  Imunossupression

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
ctable(data7$group, data7$imuno, chisq=TRUE, prop="r", useNA = "no", OR = TRUE)
```

##  Renal

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
ctable(data7$group, data7$renal, chisq=FALSE, prop="r",  useNA = "no", OR = TRUE)
```

```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
with(data7, fisher.test(renal, group))
```

## Any comorbidity

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
with(data7, ctable(group, comorbidity, prop = "r", useNA = "no", chisq = TRUE, OR = TRUE))
```

# Symptoms

```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
# Fever
data7 <-  data7 %>%
  mutate(fever = case_when(FEBRE == 1 ~ "yes",
                           FEBRE == 2 ~ "no",
                           TRUE ~ NA_character_))

# Cough
data7 <-  data7 %>%
  mutate(cough = case_when(TOSSE == 1 ~ "yes",
                           TOSSE == 2 ~ "no",
                           TRUE ~ NA_character_))

# Sore throat
data7 <-  data7 %>%
  mutate(sore_throat = case_when(GARGANTA == 1 ~ "yes",
                                 GARGANTA == 2 ~ "no",
                                 TRUE ~ NA_character_))

# Dyspnea
data7 <-  data7 %>%
  mutate(dyspnea = case_when(DISPNEIA == 1 ~ "yes",
                             DISPNEIA == 2 ~ "no",
                             TRUE ~ NA_character_))

# Respiratory discomfort
data7 <-  data7 %>%
  mutate(resp_disc = case_when(DESC_RESP == 1 ~ "yes",
                               DESC_RESP == 2 ~ "no",
                               TRUE ~ NA_character_))

# Desaturation
data7 <-  data7 %>%
  mutate(desaturation = case_when(SATURACAO == 1 ~ "yes",
                                  SATURACAO == 2 ~ "no",
                                  TRUE ~ NA_character_))

# Diarrhea
data7 <-  data7 %>%
  mutate(diarrhea = case_when(DIARREIA == 1 ~ "yes",
                              DIARREIA == 2 ~ "no",
                              TRUE ~ NA_character_))

# Vomit
data7 <-  data7 %>%
  mutate(vomit = case_when(VOMITO == 1 ~ "yes",
                           VOMITO == 2 ~ "no",
                           TRUE ~ NA_character_))

# Abdominal pain
data7 <-  data7 %>%
  mutate(abd_pain = case_when(DOR_ABD == 1 ~ "yes",
                              DOR_ABD == 2 ~ "no",
                              TRUE ~ NA_character_))

# Fatigue
data7 <-  data7 %>%
  mutate(fatigue = case_when(FADIGA == 1 ~ "yes",
                             FADIGA == 2 ~ "no",
                             TRUE ~ NA_character_))

# Olfactory loss
data7 <-  data7 %>%
  mutate(olfac_loss = case_when(PERD_OLFT == 1 ~ "yes",
                                PERD_OLFT == 2 ~ "no",
                                TRUE ~ NA_character_))

# Loss of taste
data7 <-  data7 %>%
  mutate(loss_taste = case_when(PERD_PALA == 1 ~ "yes",
                                PERD_PALA == 2 ~ "no",
                                TRUE ~ NA_character_))

# Any respiratory symptom
df <- data7 %>% 
  select(dyspnea,fatigue,desaturation,resp_disc)

soma <- function(x){
  if (sum(is.na(x))==4)
    return(NA_character_)
  else
    return(sum(!is.na(x) & x=="yes")) 
}
data7$qt_sintomas_resp_aux <- apply(df,1,soma)

data7 <-  data7 %>%
  mutate(resp_symp = case_when(qt_sintomas_resp_aux >=1 ~ "yes",
                               qt_sintomas_resp_aux ==0 ~ "no",
                               TRUE ~ NA_character_))
# Any symptom
df <- data7 %>% 
  select(dyspnea,fatigue,desaturation,resp_disc,
         fever,cough,sore_throat,diarrhea,vomit,abd_pain,olfac_loss,loss_taste)
soma <- function(x){
  if (sum(is.na(x))==12)
    return(NA_character_)
  else
    return(sum(!is.na(x) & x=="yes")) 
}
data7$qt_sintomas_aux <- apply(df,1,soma)

data7 <-  data7 %>%
  mutate(symptom = case_when(qt_sintomas_aux >= 1 ~ "yes",
                             qt_sintomas_aux == 0 ~ "no",
                             TRUE ~ NA_character_))
```

## Fever

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
with(data7, ctable(group, fever, prop = "r", useNA = "no", chisq = TRUE, OR=TRUE))
```


## Cough

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
with(data7, ctable(group, cough, prop = "r", useNA = "no", chisq = TRUE, OR=TRUE))
```


## Sore throat

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
with(data7, ctable(group, sore_throat, prop = "r", useNA = "no", chisq = TRUE, OR=TRUE))
```

##  Dyspnea

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
with(data7, ctable(group, dyspnea, prop = "r", useNA = "no", chisq = TRUE, OR=TRUE))
```

##  Respiratory discomfort

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
with(data7, ctable(group, resp_disc, prop = "r", useNA = "no", chisq = TRUE, OR=TRUE))
```

## Desaturation

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
with(data7, ctable(group, desaturation, prop = "r", useNA = "no", chisq = TRUE, OR=TRUE))
```

## Diarrhea

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
with(data7, ctable(group, diarrhea, prop = "r", useNA = "no", chisq = TRUE, OR=TRUE))
```

##  Vomit

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
with(data7, ctable(group, vomit, prop = "r", useNA = "no", chisq = TRUE, OR=TRUE))
```

## Abdominal pain

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
with(data7, ctable(group, abd_pain, prop = "r", useNA = "no", chisq = TRUE, OR=TRUE))
```

## Fatigue

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
with(data7, ctable(group, fatigue, prop = "r", useNA = "no", chisq = TRUE, OR=TRUE))
```

## Olfactory loss

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
with(data7, ctable(group, olfac_loss, prop = "r", useNA = "no", chisq = FALSE, OR=TRUE))
```

```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
with(data7, fisher.test(olfac_loss, group))
```

## Loss of taste

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
with(data7, ctable(group, loss_taste, prop = "r", useNA = "no", chisq = FALSE, OR=TRUE))
```

```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
with(data7, fisher.test(loss_taste, group))
```

## Any respiratory symptom

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
with(data7, ctable(group, resp_symp, prop = "r", useNA = "no", chisq = TRUE, OR=TRUE))
```


## Any symptom

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
with(data7, ctable(group, symptom, prop = "r", useNA = "no", chisq = FALSE, OR=TRUE))
```

```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
with(data7, fisher.test(symptom, group))
```


# Outcome

```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
# Hospital-acquired infection
data7 <-  data7 %>%
  mutate(hospital_infection = case_when(NOSOCOMIAL ==1  ~ "yes",
                             NOSOCOMIAL == 2 ~ "no",
                               TRUE ~ NA_character_))
# ICU
data7 <- data7 %>%
  mutate(icu = case_when(UTI == 1 ~ "yes",
                         UTI == 2 ~ "no",
                         TRUE ~ NA_character_))

# Intubation
data7 <- data7 %>% 
  mutate(intubation = case_when(SUPORT_VEN == 1 ~ "yes",
                                SUPORT_VEN == 2 | SUPORT_VEN == 3 ~ "no",
                                TRUE ~ NA_character_))
```

## Hospital-acquired infection

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
with(data7, ctable(group, hospital_infection, prop = "r", useNA = "no", chisq = TRUE, OR=TRUE))
```


## ICU

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
with(data7, ctable(group, icu, prop = "r", useNA = "no", chisq = TRUE, OR=TRUE))
```

### Duration of hospitalization in ICU

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
data7 <- data7 %>%
   dplyr::mutate(
    dt_entuti = as.Date(DT_ENTUTI, format = "%d/%m/%Y"),
    dt_saiduti = as.Date(DT_SAIDUTI, format = "%d/%m/%Y"),
    icu_days = as.numeric(dt_saiduti-dt_entuti)
)
```

```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
datasummary((group) ~ icu_days*(n+media+DP+mediana+minimo+maximo+q25+q75+IQR),
            data = data7, output = 'markdown')
```

```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
wilcox.test(icu_days ~ group, data = data7)
```


## Intubation

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
with(data7, ctable(group, intubation, prop = "r", useNA = "no", chisq = TRUE, OR=TRUE))
```

## Time elapsed between the start of symptoms and the date of death

We will analyze the time between the onset of the first symptoms until the patient's death. Let's create a new variable that will count the number of days of this difference.

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
# Creation of time between the start of symptoms and the date of death
data7 <- data7 %>%
   dplyr::mutate(
    dt_sin_pri = as.Date(DT_SIN_PRI, format = "%d/%m/%Y"),
    dt_evoluca = as.Date(DT_EVOLUCA, format = "%d/%m/%Y"),
    days_symp_death = as.numeric(dt_evoluca-dt_sin_pri)
)
```


```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
datasummary((group) ~ days_symp_death*(n+media+DP+mediana+minimo+maximo+q25+q75+IQR),
            data = data7, output = 'markdown')
```

```{r, echo=TRUE, eval= TRUE, message=FALSE, warning =FALSE, error=FALSE}
wilcox.test(days_symp_death ~ group, data = data7)
```


# Distribution of COVID-19 and unspecified etiologic cause by Brazilian states

The distribution of COVID-19 and unspecified etiologic cause cases by Brazilian state (SG_UF variable) is presented. 

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
with(data7, ctable(SG_UF, group, prop = "c")) #SG_UF indicates Brazilian state
```

Now we consider the rate of deaths per 100,000 live births in each group. The live births data considered is from 2019 and it is available in the link on the link: http://svs.aids.gov.br/dantps/centrais-de-conteudos/paineis-de-monitoramento/natalidade/nascidos-vivos.


```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
# Database of live births in the year 2019 in Brazil
data_birth <- read_excel("dados_nascidos_2019.xlsx")
```

## Covid-19 group

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
#Covid-19
d1 <- data7 %>% 
  filter(group == "covid-19")

valor <- data.frame(table(d1$SG_UF))
colnames(valor) <- c("uf", "n")

dt1_state <- left_join(valor, data_birth, by= "uf")

dt1_state <- dt1_state %>% 
  mutate(T1 = (n/total)*100000)

dt <- rbind(c("AC",12), c("AL",27), c("AP",16), c("AM",13), c("BA",29), 
              c("CE",23), c("DF",53), c("ES",32), c("GO",52), c("MA",21), 
              c("MT",51), c("MS",50), c("MG",31), c("PA",15), c("PB",25), 
              c("PR",41), c("PE",26), c("PI",22), c("RN",24), c("RS",43), 
              c("RJ",33), c("RO",11), c("RR",14), c("SC",42), c("SP",35), 
              c("SE",28), c("TO",17)) %>% data.table %>% `colnames<-`(c("uf","id"))
mapaUF <- readRDS("mapaUF.Rds")
dt1 <- full_join(dt, dt1_state, by = "uf")

g1 <- ggplot(dt1) + 
  geom_map(map = mapaUF, color = 'gray30', aes_string(map_id = "id", fill = "T1")) + 
  geom_path(data = mapaUF, color = 'gray30', size = .1, aes(x = long, y = lat, group = group)) + 
  theme_void() + coord_equal() + 
  labs(fill = "Death \n rate",title="COVID-19 cases") +
  scale_fill_distiller(palette="Purples",trans="reverse")  
g1
```

## Unspecified cause

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
#Unspecified cause
d2 <- data7 %>% 
  filter(group == "unspecified")

valor <- data.frame(table(d2$SG_UF))
colnames(valor) <- c("uf", "n")

dt2_state <- left_join(valor, data_birth, by= "uf")

dt2_state <- dt2_state %>% 
  mutate(T1 = (n/total)*100000)

dt <- rbind(c("AC",12), c("AL",27), c("AP",16), c("AM",13), c("BA",29), 
              c("CE",23), c("DF",53), c("ES",32), c("GO",52), c("MA",21), 
              c("MT",51), c("MS",50), c("MG",31), c("PA",15), c("PB",25), 
              c("PR",41), c("PE",26), c("PI",22), c("RN",24), c("RS",43), 
              c("RJ",33), c("RO",11), c("RR",14), c("SC",42), c("SP",35), 
              c("SE",28), c("TO",17)) %>% data.table %>% `colnames<-`(c("uf","id"))
mapaUF <- readRDS("mapaUF.Rds")
dt2 <- full_join(dt, dt2_state, by = "uf")
g2 <- ggplot(dt2) +
  geom_map(map = mapaUF, color = 'gray30', aes_string(map_id = "id", fill = "T1")) +
  geom_path(data = mapaUF, color = 'gray30', size = .1, aes(x = long, y = lat, group = group)) + 
  theme_void() + coord_equal() + 
  labs(fill = "Death rate",title="Cases with unspecified cause")  + 
  scale_fill_distiller(palette="Purples",trans="reverse") 
g2
```

## Distribution by Epidemiological Week

First, let's create the variable that will indicate the epidemiological week together with the year case.

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
data7 <- data7 %>%
  mutate(dt_sin_pri = as.Date(DT_SIN_PRI, format = "%d/%m/%Y")
)
data7 <-  data7 %>%
  mutate(seman_pri = paste(formatC(SEM_PRI, width=2, format="d", flag="0"),
                            year(dt_sin_pri),sep="/"))

d1 <- rownames_to_column(data.frame(freq(data7$seman_pri,  cum=FALSE,total=TRUE,na.last=FALSE,valid=FALSE)),var="week")
aux <- str_split(d1$week,"/",simplify=TRUE)
d1<- data.frame(aux,d1)
d1$week<- NULL
d1 <- d1 %>% arrange(X2,X1)
d1<- d1[-c(1,2),-c(5,6,7)]
d1 <- rename(d1,"Week"="X1","Year"="X2","n"="Freq","%"="X..Valid") 
```

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
#frequency table
d1
```

Now, to better understand the distribution of time, let's make a graph referring to the month of the first symptom for better visualization. The graph takes into account the percentage of unspecified cases so that the complement is the percentage of covid-19 cases.

```{r, echo=TRUE,message=FALSE,warning =FALSE,error=FALSE}
d1 <- data7[data7$dt_sin_pri < as.Date("01/04/2021",format="%d/%m/%Y"),] 
#FILTERING CASES UNTIL MARCH 2021
d1 <- d1 %>%
  mutate(month_year = paste(formatC(month(dt_sin_pri), width=2, format="d", flag="0"),
                            year(dt_sin_pri),sep="/"))

d <- prop.table(table(d1$month_year, d1$group),1)
G3 <- as.data.frame(d)
G3 <- G3[G3$Var2 == "unspecified",]
G3$Freq <- round((G3$Freq)*100, 2)
Sys.setlocale("LC_TIME","C")
G3$Var1 <- as.yearmon(G3$Var1, format = "%m/%Y") 

ggplot(data=G3, aes(x = Var1, y=Freq)) + 
  geom_line(size=1.2, color="indianred2") + 
  geom_point(size=3,color="indianred2") + 
  geom_hline(yintercept = 50) + xlab("Month of first symptom") + 
  ylab("Relative frequency(%)")
```
